Shallow EDSLs and Object-Oriented Programming

### 2016-06-22 paper/sections/Interpretations.tex
Mention multiple inheritance in S3

A well-known limitation of shallow embeddings is that they allow only a single
interpretation. Gibbons and Wu worked around this problem by accommodating
multiple interpretations using tuples and summarized how to define various types
of interpretations with a shallow embedding.
This section illustrates the OOP way of defining interpretations. Different from
their workaround, our approach is modular extensible.

\subsection{Multiple Interpretations}
%Defining additional interpretations is not trivial for shallow embedding,
%especially for functional languages.
Suppose that we want to have an additional function that checks whether a circuit is
constructed correctly. Here comes the Haskell solution:
\lstinputlisting[linerange=2-9]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
which is not modular because we add the definition of \lstinline{wellSized} by
modifying the original code. And whenever a new interpretation comes,
the arity of the tuple must be incremented and the new interpretation has to be
appended to each case.
Though Gibbons and Wu presented an modular way of defining multiple
interpretations in Haskell based on
\cite{swierstra2008data}. But such solution changes the encoding style
dramatically and may be too complex to use in practice.

Conversely, we can introduce new interpretation in a
modular and intuitive way with an OO language like Scala:
\begin{figure}
\lstinputlisting[linerange=2-15]{../src/wellsized/Circuit.scala}%APPLY:MULTIPLE_SCALA
\caption{Adding new interpretations}
\label{code:operation}
\end{figure}
Instead of modifying the original code, we define \lstinline{wellSized} modularly.
The encoding makes use of three OOP abstraction mechanisms:
\emph{inheritance}, \emph{subtyping} and \emph{type-refinement}.
Specifically, the new \lstinline{Circuit} is a subtype of
\lstinline{base.Circuit} and declares a new method \lstinline{wellSized}.
The hierarchy implements the new \lstinline{Circuit} by inheriting the corresponding class
from \lstinline{base} and
complementing the body of \lstinline{wellSized}.
Also, fields of \lstinline{Beside} are refined with the new \lstinline{Circuit}
type to avoid type mismatch in creating objects.

\subsection{Dependent Interpretations}
 \emph{Dependent interpretations} can not be defined alone which use other
 interpretations in their definition.
%Such interpretations are non-compositional.
In Haskell a dependent interpretation must be defined together with what it
dependents on and makes no exceptions on modular approaches like~\cite{}.
This prevents a new interpretation that depends on existing
interpretations from being defined modularly.
Fortunately, OO approach can still modularize such interpretations.

Before giving an example of dependent interpretations, we first show how to add new
constructs. The extended grammar of \dsl is given below,
which contains two extra constructs \emph{above} and \emph{stretch}:
\setlength{\grammarindent}{5em} % increase separation between LHS/RHS
%\lstinputlisting[linerange=39-40]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\begin{grammar}
<circuit> ::= \ldots
\alt `above' <circuit> <circuit>
\alt `stretch' <positive-numbers> <circuit>
\end{grammar}
$above\ c_1\ c_2$ combines two circuits of the same width vertically;
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns}.
Figure~\ref{code:variant} shows how to add new constructs - simply defining new classes that implement \lstinline{Circuit}.
\begin{figure}
\lstinputlisting[linerange=2-11]{../src/width/Variants.scala}%APPLY:VARIANT_SCALA
\caption{Adding new language constructs}
\label{code:variant}
\end{figure}

Not arbitrary circuits can be combined using $above$ and $stretch$
as stated in their specifications.
We hence define the \lstinline{wellSized} method for these two new constructs to
verify the constraints they imply:
\lstinputlisting[linerange=2-11]{../src/wellsized/Variants.scala}%APPLY:DEPENDENT_SCALA
Definitions of \lstinline{wellSized} for the extended cases make it a dependent interpretation, as they
rely \lstinline{width} for calculating the width of a circuit.
What distinguishes the OO approach is that \lstinline{width} and \lstinline{wellSized}
can be defined separately. Essentially, with only the signature of
\lstinline{width} exposed to the \lstinline{Circuit} we can
define \lstinline{wellSized}.

>>>>

A well-known limitation of shallow embeddings is that they allow only a single
interpretation. Gibbons and Wu worked around this problem by accommodating
multiple interpretations using tuples. However, their encoding needs to modify
the original code. Although a modular solution based on \cite{swierstra2008data}
is also presented, it complicates the encoding dramatically and may prevent pratical use.
OO approach, on the contrary, provides modular yet simple solution of defining
multiple interpretations. This section illustrates how various type of
interpretations can be defined in an OOP way.

\subsection{Multiple Interpretations}
%Defining additional interpretations is not trivial for shallow embedding,
%especially for functional languages.
Suppose that we want to have an additional function that checks whether a circuit is
constructed correctly. Here comes Gibbons and Wu's solution:
\lstinputlisting[linerange=2-9]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
which is not modular because
whenever a new interpretation is needed (e.g. \lstinline{wellSized}), the
original code has to be revised -
the arity of the tuple must be incremented and the new interpretation has to be
appended to each case.
%we add the definition of \lstinline{wellSized} by modifying the original code.

Conversely, we can introduce new interpretations in a
modular and intuitive way with an OO language like Scala, as shown in Figure~\ref{code:operation}.
\begin{figure}
\lstinputlisting[linerange=2-15]{../src/wellsized/Circuit.scala}%APPLY:MULTIPLE_SCALA
\caption{Adding new interpretations}
\label{code:operation}
\end{figure}
%Instead of modifying the original code, we define \lstinline{wellSized} modularly.
The encoding relies on three OOP abstraction mechanisms:
\emph{inheritance}, \emph{subtyping} and \emph{type-refinement}.
Specifically, the new \lstinline{Circuit} is a subtype of
\lstinline{base.Circuit} and declares a new method \lstinline{wellSized}.
The hierarchy implements the new \lstinline{Circuit} by inheriting the corresponding class
from \lstinline{base} and
complementing the body of \lstinline{wellSized}.
Also, fields of \lstinline{Beside} are refined with the new \lstinline{Circuit} type
to avoid type mismatch in creating objects.

% Multiple inheritance
We can even define \lstinline{wellSized} separately:
\begin{lstlisting}
trait Circuit { def wellSized: Boolean }
trait Id extends Circuit { ... }
...
\end{lstlisting}
And merge the two hierarchy through \emph{multiple inheritance} for providing
multiple interpretations:
\begin{lstlisting}
trait Circuit
    extends width.Circuit with wellSized.Circuit
trait Id extends Circuit
    with width.Id with wellSized.Circuit
...
\end{lstlisting}

\subsection{Dependent Interpretations}
 \emph{Dependent interpretations} use other interpretations in their definition.
%Such interpretations are non-compositional.
In Haskell a dependent interpretation must be defined together with what it
dependents on and makes no exceptions on modular approaches like~\cite{}.
This prevents a new interpretation that depends on existing
interpretations from being defined modularly.
Fortunately, OO approach does not have such restriction.

Before giving an example of dependent interpretations, we first show how to add new
constructs in OO approach. The extended grammar of \dsl is given below,
which contains two extra constructs \emph{above} and \emph{stretch}:
\setlength{\grammarindent}{5em} % increase separation between LHS/RHS
%\lstinputlisting[linerange=39-40]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\begin{grammar}
<circuit> ::= \ldots
\alt `above' <circuit> <circuit>
\alt `stretch' <positive-numbers> <circuit>
\end{grammar}
$above\ c_1\ c_2$ combines two circuits of the same width vertically;
\emph{stretch ns c} inserts additional wires into the circuit \emph{c} by
summing up \emph{ns}.
Figure~\ref{code:variant} shows how to add them to \dsl - simply defining new classes that implement \lstinline{Circuit}.
\begin{figure}
\lstinputlisting[linerange=2-11]{../src/width/Variants.scala}%APPLY:VARIANT_SCALA
\caption{Adding new language constructs}
\label{code:variant}
\end{figure}

From the specifications we can see that not arbitrary circuits can be combined using $above$ and $stretch$.
We hence define the \lstinline{wellSized} method for them to
verify the constraints they imply:
\lstinputlisting[linerange=2-11]{../src/wellsized/Variants.scala}%APPLY:DEPENDENT_SCALA
Definitions of \lstinline{wellSized} for the extended cases make it a dependent
interpretation, as they can not be defined without \lstinline{width}.
Note that \lstinline{width} and \lstinline{wellSized} are defined separately.
Essentially, it is sufficient to define \lstinline{wellSized} with only the signature of \lstinline{width} in \lstinline{Circuit}.
